Influence of Time Integration Methodology on Double Detonation Type Ia Supernova Nucleosynthesis

### 2023-06-20 subch_tols/paper.tex
add more figs
update

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We explore the early evolution of flame ignition and spreading on the
surface of a neutron star in three-dimensions, in the context of X-ray
bursts.  We look at the nucleosynthesis and morphology of the burning
front and compare to two-dimensional axisymmetric simulations to gauge
how important a full three-dimensional treatment of the flame is for
the early dynamics.  Finally, we discuss the progress toward full-star
resolved flame simulations.\end{abstract}

\keywords{convection---hydrodynamics---methods: numerical---stars: neutron---X-rays: bursts}

@@

@@

Integration of reaction networks takes various forms in astrophysical
simulation codes.

@@

All simulations use the same network, a 22 nucleus network designed for
He and C burning, using the ReacLib rates \cite{reaclib} and produced with
\pynucastro \citep{pynucastro2}.  This network, called {\tt subch\_simple} was
described in \cite{zhi2023}.

The reaction system is integrated using a version of the VODE ODE integrator
\citep{vode} ported to C++.  Our modifications to this integrator are described
in \citet{castro_simple_sdc}.

If during the advance of a timestep an error is generated (negative
density, ODE integration fails, mass fractions don't sum to 1), the
step is thrown-out and retied with a smaller timestep.  This is done
on a level-by-level basis in the overall AMR subcycling hierarchy.

We will run a suite of simulations all using the same initial model.
The initial model is constructed following the methodology in
\citet{subchandra}.  We use an isothermal core of C/O of $1.1~M_\odot$
with $T = 10^7~\mathrm{K}$ and a thin transition region where the
temperature ramps up to $1.75\times 10^8~\mathrm{K}$ and the
composition changes to 99\% \isot{He}{4} and 1\% \isot{N}{14}.  This
added \isot{N}{14} helps produce protons via
$\isotm{N}{14}(\alpha,\gamma)\isotm{F}{18}(\alpha,p)\isotm{Ne}{21}$
which then allow for
$\isotm{C}{12}(p,\gamma)\isotm{N}{13}(\alpha,p)\isotm{O}{16}$.  This
sequence can be faster than
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$, as pointed out by
\citet{shenbildsten} and is important for getting the detonation speed
correctly.  This He envelope is then integrated isentropically.  The
density of the transition from the underlying CO white dwarf to the He
envelope was selected to yield an envelope mass of $0.05~M_\odot$.

@@

@@

@@

Show cost plot

@@

The ignition of the carbon detonation remains numerically challenging for the ODE integrator, especially with large timesteps.  \MarginPar{add some info on why it fails}
In the near future we will show how to include various nuclear
statistical equilibrium approximations into the simplified-SDC
formalism while retaining second-order accuracy.

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We explore the coupling of hydrodynamics and reactions in simulations of
the double detonation model for Type Ia supernovae.
\end{abstract}

\keywords{convection---hydrodynamics---methods: numerical}

@@

These events are modeled using the Euler equations with reactive and
gravitational sources:
\begin{eqnarray*}
\ddt{\rho} + \nabla \cdot (\rho \Ub) &=& 0 \\
\ddt{(\rho \Ub)} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &=& \rho \gb \\
\ddt{(\rho E)} + \nabla \cdot (\rho \Ub E + \Ub p) &=& \rho \Ub \cdot \gb + \rho \dot{S} \\
\ddt{(\rho X_k)} + \nabla \cdot (\rho \Ub X_k) &=& \rho \omegadot_k
\end{eqnarray*}

Integration of reaction networks takes various forms in astrophysical
simulation codes.  The most common method is via operator splitting:
the advection and reactive terms are treated independently, with each
process working on the result of the other.  Often, Strang-splitting
\citep{strang:1968} is used, which alternates advection and reaction
to yield second-order accuracy in time.

@@

@@

Our default CFL number is 0.2 for these simulations, and the CFL constraint
is the only timestep constraint used.  For larger CFL numbers, the ODE integration
sometimes encounters problems \MarginPar{more}.

All simulations use the same network, the {\tt subch\_simple} as
described in \cite{zhi2023}.  This includes 22 nuclei and 94 rates
(from ReacLib \citealt{reaclib}) and is produced with \pynucastro
\citep{pynucastro2}.  Importantly, we include
$\isotm{N}{14}(\alpha,\gamma)\isotm{F}{18}(\alpha,p)\isotm{Ne}{21}$
which creates protons to allow for
$\isotm{C}{12}(p,\gamma)\isotm{N}{13}(\alpha,p)\isotm{O}{16}$.  This
sequence can be faster than
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$, as pointed out by
\citet{shenbildsten} and is important for getting the detonation speed
correctly.  To reduce the size of the network, some
$(\alpha,p)(p,\gamma)$ and $(\alpha,\gamma)$ rates are combined into
an effective $(\alpha,\gamma)$ rate.  Screening is provided following
the procedure in \citet{wallace:1982}, combining the screening
functions of \citet{graboske:1973,alastuey:1978,itoh:1979}.

The reaction system is integrated using a version of the VODE ODE integrator
\citep{vode} ported to C++.  Our modifications to this integrator are described
in \citet{castro_simple_sdc}.  We use an analytic approximation to the Jacobian
and found that we get best results when we disable the caching of the Jacobian
in the integrator.

If during the advance of a timestep an error is generated (negative
density, ODE integration fails, mass fractions don't sum to 1, CFL constraint violated at the new time), the
step is thrown-out and retied with a smaller timestep.  This is done
on a level-by-level basis in the overall AMR subcycling hierarchy.

We will run a suite of simulations all using the same initial model.
The initial model is constructed following the methodology in
\citet{subchandra}.  We use an isothermal core of C/O of $1.1~M_\odot$
with $T = 10^7~\mathrm{K}$ and a thin transition region where the
temperature ramps up to $1.75\times 10^8~\mathrm{K}$ and the
composition changes to 99\% \isot{He}{4} and 1\% \isot{N}{14}.    This He envelope is then integrated isentropically.  The
density of the transition from the underlying CO white dwarf to the He
envelope was selected to yield an envelope mass of $0.05~M_\odot$.

@@

Figure~\ref{fig:lap_rho_sequence} shows a Schlieren-style plot, which
highlights density gradients.  It is constructed by plotting
$\log_{10}(|\rho^{-1}\nabla^2\rho|)$.  This clearly shows the
compression wave launched by the He detonation propagating in toward
the center of the white dwarf.  The middle row shows this compression
wave converging slightly off-center which then leads to the ignition
of the C detonation.  This also highlights how the speed of the He
detonation is important in determining how off-center the C detonation
ignition will be.  Our choice of network was motivated by ensuring
that we capture the energy release from He burning accurately, in
particular the inclusion of the
$\isotm{C}{12}(p,\gamma)\isotm{N}{13}(\alpha,p)\isotm{O}{16}$
sequence.

@@

\begin{figure}[t]
\centering
\plotone{subch_ni56}
\caption{\label{fig:ni56} \isot{Ni}{56} mass vs.\ time for the different simulations.}
\end{figure}

\begin{figure}[t]
\centering
\plotone{subch_cpu}
\caption{\label{fig:cpu} MPI-hours of GPU time for the different simulations.}
\end{figure}

@@

Figure~\ref{fig:cpu} shows the computational expense of the
simulations, in terms of node hours.  All simulations were run on 4
nodes (32 AMD GPUs total).  We note that we have not spent any time on
load-balancing these simulations, so we should consider this to be
only a guide.  At the start of the simulation, there are 56 boxes on
the finest grid (representing 0.77\% of the domain) and at 1 s of
evolution there is are 60 boxes covering 5.64\% of the domain.  The
figure shows that the Strang simulation with a CFL of 0.2 is the
cheapest, but as we saw above, this does not get the nucleosynthesis
correct.  All of the other Strang simulations are more costly than the
baseline SDC simulation.  This shows that not only does the
simplified-SDC simulation accurately evolve the nucleosynthesis, it
does so in a very efficient fashion.

@@

The ignition of the carbon detonation remains numerically challenging
for the ODE integrator, especially with large
timesteps.  \MarginPar{add some info on why it fails} In the near
future we will show how to include various nuclear statistical
equilibrium approximations into the simplified-SDC formalism while
retaining second-order accuracy.  We will also explore different
integrators and auto-differentiation for the Jacobian (in particular,
the compositional contribution from the screening)
